---
title: "R Basics"
author: David J. Stokes | Teaching Coordinator | Data Science Academy
format: 
  html:
    theme: cerulean
    toc: true
editor: visual
---

# R Data Types & Structure

-   Let's begin with some introductory R programming concepts; data types and data structures.

-   As we move through this lesson, run the code in the code chunks, observe and consider the output, and complete the tasks.

## Data Types

-   Let's discuss a few R data types that are typical in the data science programming process:

### Numeric

-   this is the default quantitative data type in R
    -   Examples include: `4`, `pi`, `10e-4`, `2423.2`

> NOTE: You can assign a value to an object/variable that you can later reference (or call) by name using the arrow `<-`

```{r}
# (in R, variable names are flexible but are case sensitive)
num1 <- 4 # Here I'm storing the value 4 in an object called num1
num.2 <- pi
num_3 <- 10e-4
Num4 <- 2423.2
```

> NOTE: You can perform numerical operations on numerical values (see the examples below)

```{r}
num1 + num.2 #4 + pi

num_3*Num4 #10e-4*2423.2
# NOTE: 10e-4 is scientific notation for 10*10^(-4) or 10^-3
```

**📊 Task 1.4**

-   In the code chunk below:

1)  multiply `num_3` by `10^3` and display the result.

2)  try another calculation using at least three of the variables `num1`, `num.2`, `num_3`, or `Num4`, and display the results.

```{r}
# [Your code and comments go in this code chunk] 
# (Hint: it can be useful to copy and paste)

```

### Character

-   this is the default data type for strings and text data
    -   Examples include: `"words & phrases"`, `"RED"`, `"756"`

```{r}

charA <- "Hello"
charB <- "I am"
charC <- "[your name goes here]"

## In R, you can pass objects to a function. Here is an example using the paste() function, with charA, charB, and charC
paste(charA, charB, charC)
```

**📊 Task 1.5**

1)  Run the code below:

```{r}
a <- "1"
b <- "2"

a
b
```

2)  in the code chunk below, try a + b and describe what happens and why you think this happened.

```{r}
#[Your code and comments go in this code chunk] 
```

**\[Your answer goes here\]**

3)  go back to the code chunk above and comment out the line where you added `a` and `b`.

### Logical

-   this data type represents boolean values
    -   Example include: `TRUE` and `FALSE`

```{r}
# Logicals tell us if the conditions are met

2 + 2 == 4 

7 < 6 

```

## Data Structures

### Vectors

-   Vectors are one-dimensional data structures that contain the same type. For example, you can have a numerical vector consisting of numeric data types or a character vector consisting of character data types.

```{r}

# the c() function "combines" elements (separated by commas) into a vector.
even_vec <- c(2,4,6,8,10)

odd_vec <- even_vec - 1

#NOTE: you can perform operations to each element of a vector (vectorization) without needing to iterate through each element.
```

**📊 Task 1.6**

1)  In the code chunk below, create one numeric vector and one character vector, each consisting of at least two components and name & store them in new objects.

```{r}
#[Your code and comments go in this code chunk]

```

2)  In the code chunk below, perform a calculation on your numeric vector that transforms the elements of the vector and run the code to display the results.

```{r}
#[Your code and comments go in this code chunk]
```

3)  In the code chunk below, access and display the second element of your character vector using the following bracket notation `vectorName[element position]` (e.g., even_vec\[2\]).

```{r}
#[Your code and comments go in this code chunk]

```

4)  Use the c() function to combine the second element of your character vector with the second element of your numeric vector, then name & save this as a new object.
    -   Run the code and display the results of your new object.
    -   Describe the new object in the space below the code (i.e., what type of vector is this new object?).

```{r}
#[Your code and comments go in this code chunk]

#HINT-example NewVec <- c(OldNumVec[2], OldCharVec[2])

```

**\[Your answer goes here\]**

### Matrices

-   Matrices are two-dimensional R objects that contain data of the same type.
    -   Like vectors, matrices can be character or numeric, but not both.
-   Let's build some example matrices

```{r}

# numeric matrix.. Notice that the option "byrow" has been set to TRUE
frst_mtrx <- matrix(c(1,2,3,
                      4,5,6,
                      7,8,9,
                      10,11,12), nrow = 4, ncol = 3, byrow = TRUE)

char.mat <- matrix(c("I","I",
                     "Love", "Love",
                     "Double", "Double",
                     "Chex", "Chex"), nrow = 4, ncol = 2, byrow=TRUE)


```

**📊 Task 1.7**

-   In the code chunk below:

1)  modify `frst_mtrx` by deleting the `byrow` option and save this as a new object called `scnd_mtrx`.

2)  use the `colSums()` function on `scnd_mtrx`, run the code to display the results, and describe what this function does (in the space below the code chunk).

```{r}
#[Your code and comments go in this code chunk] 

```

**\[Your answer goes here\]**

3)  What do you think would happen to the matrix type of `frst_mtrx` if we replaced a row in `frst_mtrx` with a row of `char.mat`? (HINT: what happened to the vector type when we added a character value to our numeric vector?)

**\[Your answer goes here\]**

4)  You can access components of a matrix (or other two dimensional R object, like Tibbles and Dataframes) using bracket notation, e.g., `MatrixName[row location, column location]`.

-   To access an entire row, you can use `MatrixName[row location, ]`
-   To access an entire column, you can use `MatrixName[ ,column location]`

In the code chunk below, display the first row of `frst_mtrx`, and then multiply this row by 2 and show the results.

```{r}
#[Your code and comments go in this code chunk]
```

### Dataframes & Tibbles

-   Finally, we've arrived at the data structures for data science. Dataframes and Tibbles are two-dimensional data structures that can contain different data types in different columns. Typical dataset processing will be done using these objects (e.g., subsetting, visualizations, modeling, etc.)

-   Tibbles are structures in the `tidyverse` (a collection of useful packages) and were created to be upgrades to dataframes, but are otherwise very similar.

-   To explore Tibbles we will read in a dataset that will have this structure by default.

**📊 Task 1.8**

1)  run the code below to read in the 'College.csv' data

```{r, message=FALSE}
library(readr) #included in the tidyverse
College <- read_csv("College.csv")
```

2)  In the code chunk below, use the `dim()` function and the `colnames()` function on the College dataset.
    -   Run the results, and below your code describe what these functions do. (HINT: you can type `?dim`, or `?colnames` to bring up the R documentation on these functions in the Help window).

```{r}
#[Your code and comments go in this code chunk]

```

**\[Your answer goes here\]**

3)  In the code chunk below use the `head()` function on the College dataset, run the code, and below your code describe the information that is given.

```{r}
#[Your code and comments go in this code chunk]

```

**\[Your answer goes here\]**

4)  Finally, use the `as.data.frame()` function on the College dataset and store the results in an object called `df_college`.
    -   Run the `str()` function on `df_college`, display and describe the results. (You may also try the other functions on `df_college` for comparison).

```{r}
#[Your code and comments go in this code chunk]

```

**\[Your answer goes here\]**

| 
| 
| 
| 
