---
title: "Visualizations in Base R"
author: David J. Stokes | Teaching Coordinator | Data Science Academy
format: 
  html:
    theme: cerulean
    toc: true
editor: visual
---

# Visualizations in (base) R

## Review

-   Let's recall concepts from Day 1

-   💬 Take 5 minutes: Find a neighbor and discuss concepts from yesterday that you learned or were able to review. Please consider:

    -   Did you find anything insightful?
    -   Are there any concepts that you'd want more application context for?
    -   Discuss the dataset that you chose with your partner and how your data dictionary development is going.

-   After discussing with your neighbor, please consider sharing some thoughts with the entire group.

## Today

-   Today, you will be able to
    -   Prepare data for visualizations
        -   write user-defined functions
    -   Use ggplot2 for data visualizations
    -   Develop a data investigation/research question of interest
        -   in consideration of the data you have chosen
-   Ok, let's get started...

**📊 Task 2.1**

-   Run the code below to load the needed R packages

```{r, message=FALSE}
## loading required packages 

library(tidyverse) 

# includes ggplot2, dplyr, and readr [e.g., library(ggplot2); library(dplyr); library(readr)]

```

**📊 Task 2.2**

-   Run the code below to load in the datasets of interest
-   NOTE: You may need to change your filepath or set your `working directory`

```{r, message=FALSE}

Fish6Species <- read_csv("Fish6Species.csv")
RisingGlobalTemp <- read_csv("RisingGlobalTemp.csv")

```

## Investigations and Visualizations (base R)

*Rising Global Temperatures dataset*

-   This dataset has 3 variables (note the first column is just an index, again). The "gistemp" variable represents average temperature differences from the 1986-2005 average, for each year from 1880 to 2018. "year" represents the year in which the temperatures were taken. For example, a measurement of 0.5 for a given year would represent an average temperature of 0.5 degrees Celsius warmer than the average temperature from 1986-2005.

### The basic plot function

-   The R base plot function is pretty powerful. The syntax is simple and the defaults can be useful in visualizing trends.

-   Let's take a look at the basic plot function for the quantitative variables in the Rising Global Temperatures (`RisingGlobalTemp`) dataset.

Let's look at the column names so that we can grab the variables of interest by name.

```{r}
## run this to get the column names
colnames(RisingGlobalTemp)
```

-   ❓ What will happen if we use the basic plot function on one *quantitative* variable? (Also...recall the Dataset\$Variable syntax).

```{r}
plot(RisingGlobalTemp$gistemp) #scatterplot (by index)
```

-   ❓ What about for two quantitative variables?

```{r}
plot(x = RisingGlobalTemp$year, y = RisingGlobalTemp$gistemp) # bivariate plot
```

**📊 Task 2.3**

1)  In the space below, describe the difference between the one variable plot and two variable plot.

**\[Your answer goes here\]**

-   ❓ What else can we add to the base plot function? (colors, overlays, linetypes... the list goes on)

```{r}
plot(RisingGlobalTemp$year,RisingGlobalTemp$gistemp, xlab="Temperature Difference",
     ylab="Year", col="purple") # bivariate plot
## the abline adds a horizontal line at y=0. 
## lty changes the line type. lwd changes the line width.
abline(h=0, col="orange", lty=3, lwd=2)
```

**📊 Task 2.4**

1)  Note, there is a mistake with the axis labels in the graph above. In the codechunk below do the following:

-   fix the axis labels,
-   change the point color to red, and
-   change the line type to a different format.

```{r}
## #[Your code and comments go in this code chunk] HINT: You may want to copy and paste the above code and modify accordingly.

```

2)  In the space below, describe what is happening in the graph. Use the information given about the dataset to interpret the visualization.
    -   Do you have any hypotheses or additional information from prior knowledge about what appears in the graph?

**\[Your answer goes here\]**

-   ❓ What does the basic plot function do for categorical variables (i.e., what is the default output)?

**Fish - Six Species dataset**

-   To investigate this, let's look at the Fish - Six Species dataset (`Fish6Species`).

```{r}
## run this to get the colnames
colnames(Fish6Species)
```

-   This dataset contains 12 variables (NOTE: the variables "...1" and "X" are just indices, so let's drop them from the dataset).

```{r}
# base R version
# Fish6Species <- Fish6Species[,-c(1,2)] 

# dplyr (tidyverse) version

Fish6Species <-
  Fish6Species %>%
  select(-`...1`,-X)

# NOTE: the minus (-) signs drop the variables from the data
  
```

**📊 Task 2.5**

1)  In the code chunk below, run a function to get the dimensions of the new Fish6Species dataset, and list the new number of variables below your code (in the code chunk) as a `comment`.

```{r}
#[Your code and comments go in this code chunk]

```

2)  In the code chunk below, write code to display the names of the variables.

-   In the space below the code, write whether we got the data subset of interest and how you know whether we did or not.

```{r}
#[Your code and comments go in this code chunk]

```

**\[Your answer goes here\]**

### Back to plotting

-   ❓ What happens if you use the basic plot function on one *categorical* variable? (NOTE: we read the non-quantitative variables in as `characters`... so we need to change them to `factors`.

```{r}
# change the variables Species and Region into factors, and overwrite
Fish6Species <- 
  Fish6Species %>%
  mutate(Species = as.factor(Species), 
         Region = as.factor(Region))

summary(Fish6Species[,2:3])
```

```{r}
## run this to view the default output for one categorical variable
plot(Fish6Species$Species)
```

-   💬 What are your thoughts about the display?

-   ❓ Two categorical variables?

```{r}
plot(Fish6Species$Species, Fish6Species$Region) ## Mosaic plot
```

-   The default output from the base plot function for two categorical variables is a mosaic plot.

-   A mosaic plot is usually very informative and can be thought of as a visual representation of a two by two frequency table. By observing mosaic plots, trends are often discernible.

-   In order to make more sense out of the plot, let's look at the levels

```{r}
levels(Fish6Species$Species)
levels(Fish6Species$Region)
```

-   Still, with this information, the mosaic plot is not very informative.

-   Attempting to modify the axes and add colors to the mosaic plot is not easily done within the base R plot function. So, let's consider some other packages.

**📊 Task 2.6**

1)  Individually, or in collaboration with your chosen project group:

-   load in your chosen dataset
-   use the base plot function to generate at least two plots
    -   add titles, color, and axis labels where possible

```{r}

#[Your code and comments go in this code chunk]

```

2)  In the space below, interpret the plots that you've generated.

**\[Your answer goes here\]**
